---
title: "Red Wine Quality Analysis"
output: html_document
---
```{r echo=FALSE, message=FALSE, warning=FALSE}
wq <- read.csv('wineQualityReds.csv')
library(ggplot2)
```

#Univariate Analysis
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wq[, -1])
```

Median quality is 6. Density doesn't vary much (min = 0.99, max =1.00). 
Maximum residual sugar level is much higher than Mean and 3rd Quantile, so it may
have outliers. Similarly, chlorides and total.sulfur.dioxide seemt to have 
outliers.


###Distribution of quality
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality), data = wq) +
  geom_histogram(binwidth = 1, color ="black", fill = "white")
```

Mean of quality is 5.6, median is 6, and 75% percentile is 6. 
Most red wines are rated between 5 and 7. Only a very few number of wines are 
outside 5 and 7. Considering that # of wines with rating 8 is so small, 
it may not be very meaningful to look at the data associated with quality 8. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity), data = wq) +
  geom_histogram(color ="black", fill = "white") 
```

Mean of volatile acidity is 0.53, median is 0.52, and 75% percentile is 0.64. Its distribution is 
slightly skewed right.  

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol), data = wq) +
  geom_histogram(color ="black", fill = "white") 
```

Mean of alcohol is 10.4, median is 10.2, and 75% percentile is 11.1. Its distribution is skewed right.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = sulphates), data = wq) +
  geom_histogram(color ="black", fill = "white") 
```

Mean of sulphates is 0.66, median is 0.62, and 75% percentile is 0.73. Some samples with value 2
are outliers.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid), data = wq) +
  geom_histogram(color ="black", fill = "white") 
```

Mean of citric.acid is 0.27, median is 0.26, and 75% percentile is 0.42.
Many wines don't have citric.acid at all.

#Bivariate Analysis
### Correlation between quality and other metrics
```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(wq[, -1])
```

###Plot of the correlation matrix
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(GGally)
ggcorr(wq[, -1],  size = 3, max_size = 9) + ggtitle("Correlation matrix")
```

Based on the correlation analysis, alcohol, volatile.acidity, sulphates, 
and citric.acid are strongly correlated with wine qualtiy. 

There are several chemical properties that are strongly correlated each other: 
pH, fixed.acidity, citric.acid

I wanted to see how the scatter plot of quality vs strongly correlated 
properties.

### Scatter plots of most correlated chemical properties vs quality

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity, y= quality), data = wq) + geom_jitter(alpha = 0.3)
```

Correlation between volatile acidity and wine quality is -0.39, and this scatter plots confirms the negative correlation. As the volatile acidity decreases, wine quality improves even though the same 
quality wine has various acidity levels. For the highest quality wines 
(quality 7 and 8), acdity level is mostly below 0.7. So, this implies that 
too high level of acidity level gives a bad taste (as explained in the document), 
and such wines won't get a good qulity score.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y= quality), data = wq) + geom_jitter(alpha = 0.3)
```

Correlation between alcohol and wine quality is 0.48. As the alcohol level increases, wine quality improves. For the highest quality 
wines (quality 7 and 8), alcohol levels are mostly over 10. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = sulphates, y= quality), data = wq) + geom_jitter(alpha = 0.3)
```

Correlation between sulphates and wine quality is 0.25. As the sulphates level increase, wine quality improves. For the very high 
quality wines, sulphates levels are mostly over 0.5.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid, y= quality), data = wq) + geom_jitter(alpha = 0.3)
```

citric.acid has a relatively strong correlation with quality (0.23). 
The document explains that it can add "freshness", but based on this plot, 
it is not clear whether citric.acid really affects the wine quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid, y= volatile.acidity), data = wq) + 
  geom_jitter(alpha = 0.3) + geom_smooth(method = 'lm', color = 'red')
```

Correlation between citric acid and volatile acidity is -0.55. As citric.acid increases, volatile.acidity decreases. 

###Box plot: quality vs volatile.acidity
Tring to show the characteristics of chemical properties for a given quality. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = factor(quality), y= volatile.acidity), data = wq) + 
  geom_boxplot() + geom_jitter(alpha=0.3)
```

Mean volatile.acidity of qulaity 7 and 8 wines is around 0.4 
while mean volatile.acidity of qulaity 3 wine is around 0.8.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = factor(quality), y= alcohol), data = wq) + geom_boxplot() +
  geom_jitter(alpha=0.3)
```

The mean alcohol level of quality 7 and 8 wines is close to 12 while the mean alcohol level of 
quality 3 and 4 wines is close to 10. It seems that alcohol levels of quality 3 and 4 wines 
are higher than that of qulity 5 wine, but it is probably due to the lack of samples
in quality 3 and 4 wines. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = factor(quality) , y= sulphates), data = wq) + geom_boxplot() +
  geom_jitter(alpha=0.3)
```

The mean sulphates level of quality 7 and 8 wines is close to 0.75 while the mean alcohol level of 
quality 3 and 4 wines is close to 0.6. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = factor(quality) , y= citric.acid), data = wq) + geom_boxplot() +
  geom_jitter(alpha=0.3)
```

Correlation between citric.acid and qualtiy is 0.23. The mean citric acid level of quality 7 and 8 wines is close to 0.35 while the mean alcohol level of 
quality 3 and 4 wines is close to 0.25. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = factor(quality) , y= density), data = wq) + geom_boxplot() +
  geom_jitter(alpha=0.3)
```

Correlation between density and qualtiy is -0.175. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = factor(quality) , y= residual.sugar), data = wq) + geom_boxplot() 
```

Correlation between residual sugar and qualtiy is 0.013. 
The residual sugar level doesn't vary much for different quality wines.

###Box plot: qulity vs total.sulfur.dioxide when free.sulfur.dioxide>50
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = factor(quality) , y= total.sulfur.dioxide), data = 
subset(wq, free.sulfur.dioxide>50)) + geom_boxplot() 
```

The document that explains wine data indicates that total.sulfur.dioxide affects 
the taste of wine if free.sulfur.dioxide is over 50ppm. Based on this box plot, 
it seems that better quality wines have lower level of total.sulfur.dioxide, 
but it is hard to make a conclusion because # of samples is small (16 samples 
over 50ppm)

#Multivariate Analysis
### volatile.acidity, alcohol vs quality
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(RColorBrewer)
wq$alcohol.disc <- round(wq$alcohol)
ggplot(aes(x = factor(quality) , y=volatile.acidity, fill = factor(alcohol.disc)), 
       data = wq) + geom_boxplot() + scale_fill_brewer()
```

For any given alcohol level, higher quality wines have lower volatile acidity levels.
For example, if alcohol.disc = 9, mean volatile acidity level of quality 5 wine 
is 0.6, and mean volatile acidity level of quality 7 wine is close to 0.4.

### volatile.acidity, alcohol vs quality
```{r echo=FALSE, message=FALSE, warning=FALSE}
wq$volatile.acidity.disc <- round(wq$volatile.acidity*5)/5
wq$volatile.acidity.disc <- pmin(1,wq$volatile.acidity.disc)
wq$volatile.acidity.disc <- pmax(0.2,wq$volatile.acidity.disc)
ggplot(aes(x = factor(quality), y=alcohol, fill = factor(volatile.acidity.disc)), 
       data = wq) + geom_boxplot() + scale_fill_brewer()
```

Limited the range of acidity level between 0.2 and 1
Discretize the alcohol level to understand volatile.acdity vs quality for the 
given alcohol level. 
For any given volatile.acidity, higher quality wines have higher alcohol levels.
For example, if volatile.acidity.disc = 0.2, mean alcohol level of quality 5 wine 
is 10.5, and mean alcohol level of quality 7 wine is close to 11.5. This positive
correlation betwee alcohol level and quality holds for any given volatile.acidity.

###Linear model

```{r echo=FALSE, message=FALSE, warning=FALSE}
m1 <- lm(quality ~ alcohol , data = wq )
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + sulphates)
m4 <- update(m3, ~ . + citric.acid)
m5 <- update(m4, ~ . + total.sulfur.dioxide)
library(memisc)
mtable(m1, m2, m3, m4, m5)
```

citric.acid and total.sulfur.dioxide don't imporve R-squred. The R-squared is 
not high enough to explain the variation in the quality, but considering that 
the quality is a very subjective metric depending on people's tastes, this 
number may not be too bad. 

#Final Plots and Summary

###Plot 1

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality), data = wq) +
  geom_histogram(binwidth = 1, color ="black", fill = "white") +
  ggtitle("Histogram of wine quality") +
  xlab("Quality") + ylab("Count") 
```

This plot shows the distribution of wine quality. Considering that the main objective of the analysis is to find the chemical properties that affects wine quality, it is important to know how the wine quality distributes. 
Mean of wine quality is 5.6, median is 6, and 75% percentile is 6. 
Most red wines are rated between 5 and 7. Only a very few number of wines are 
outside 5 and 7. 

###Plot 2

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(GGally)
ggcorr(wq[, -1],  size = 3, max_size = 9) + ggtitle("Correlation matrix")
```

Because there are many chemical properties in the data set, it is necessary to look at the correlation matrix and understand which properites are more correlated with quality, or with other other chemical properties. Based on this plot, alcohol, volatile.acidity, sulphates, 
and citric.acid are strongly correlated with wine qualtiy. 

There are several chemical properties that are strongly correlated each other: 
pH, fixed.acidity, citric.acid


###Plot 3
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = factor(quality) , y=volatile.acidity, fill = factor(alcohol.disc)), 
       data = wq) + geom_boxplot() + scale_fill_brewer() +
  ggtitle("Quality by Volatile acidity and Alcohol") +
  ylab("Volatile acidity (g/dm^3)") +
  xlab("Quality") +
  labs(color = "Alcohol")
```

I have seen that alcohol level and volatile acidity are strongly correlated with 
winq quality, but it is necessary to plot them together to understand how they 
are correlated with wine quality together. Based on this plot, for any given 
alcohol level, higher quality wines have lower volatile acidity levels.
For example, if alcohol.disc = 9, mean volatile acidity level of quality 5 wine 
is 0.6, and mean volatile acidity level of quality 7 wine is close to 0.4.

#Reflection

This wine data set includes 1599 samples with 12 chemical properties and one 
quality score. The main purpose of this analysis is to find out which chemical 
properties detemines the quality of wines.

I first studied individual variables to understand how they distributed. Then, 
I looked at correlations between chemical properties and quality. Based on the 
correlations, volatile acid and alcohol levels are strongly correlated with 
qualtiy, so I started looking into more details. 

Based on the study, it is clear that volatile acid and alcohol levels affects 
the quality of wines, and people willl get different tastes. However, R-squared
from the liner model was not high enough to explain all the variations in the 
quality, which means that people will give different scores for different 
reasons. Considering that wine quality is a very subjective metric, I also 
didn't expect to see very high R-squred, but I think it will be interesting 
to get other properties that may affect wine qualities. It may also depend 
on people who judge the qualities because each person has different tastes, 
or preferences. Nonetheless, I think this analysis can identify the most 
important chemical properties that determine the quality of wines, which can 
help consumers or manufatures to select or make better wines.

It was hard to find meanings from the multi-variate analysis (plots) due to over-plotting of samples. After changing the plot to a box-plot as suggested, I could draw meaningful conclusions from the multivarate plots (volatile acidity, alcohol vs quality). From this experience,
I could understand the importance of better visualizations of data